---
title: "Assignment 10: Data Scraping"
author: "Yimei Du"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r 1. Project Setup, message = FALSE}
#1 
#Install familiar packages
library(tidyverse);library(lubridate);library(viridis);library(here)
here()

#install.packages("rvest")
library(rvest)

#install.packages("dataRetrieval")
library(dataRetrieval)

#install.packages("tidycensus")
library(tidycensus)

# Set theme
mytheme <- theme_gray() +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)

here()

```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2024 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024>
 
Indicate this website as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2 
URL <- read_html('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024')
class(URL)
```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PWSID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".

```{r scrape.the.data}
#3 
name <- URL %>% 
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% 
  html_text()
name

id <- URL %>% 
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>% 
  html_text()
id

ownership <- URL %>% 
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% 
  html_text()
ownership

MGD <- URL %>% 
  html_nodes("th~ td+ td, th~ td+ td") %>% 
  html_text()
MGD



```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... 

5. Create a line plot of the maximum daily withdrawals across the months for 2024, making sure, the months are presented in proper sequence.

```{r create.a.dataframe.from.scraped.data}
#4 
df_durham <- data.frame("Month" =
                          c("Jan","Feb","Mar","Apr","May","Jun","Jul","Aug","Sept","Oct","Nov","Dec"), 
                             "Year" = rep(2024,12),
                             "Maximum_Day_Use" = as.numeric(MGD))

df_durham <- df_durham %>% 
  mutate(
    Date = my(paste(Month,Year))) %>% 
  arrange(Date)
  

#5 
ggplot(df_durham,aes(x=Date,y=Maximum_Day_Use)) + 
  geom_line() + 
  geom_smooth(method="loess",se=FALSE) +
  labs(title = paste("Maximum Day Use (MGD) in 2024"),
       subtitle = "DUrham Municipal LWSP",
       y="Maximum Day Use (MGD)",
       x="Date")

```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function with two inputs - "PWSID" and "year" - that:
  - Creates a URL pointing to the LWSP for that PWSID for the given year
  - Creates a website object and scrapes the data from that object (just as you did above)
  - Constructs a dataframe from the scraped data, mostly as you did above, but includes the PWSID and year provided as function inputs in the dataframe. 
  - Returns the dataframe as the function's output


```{r construct.a.scraping.function}
#6. 
scrape.it <- function(year, id){

the_durham_url <- read_html(
  paste0('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=', id, '&year=', year)
)

name <- 'div+ table tr:nth-child(1) td:nth-child(2)'
id <- 'td tr:nth-child(1) td:nth-child(5)'
ownership <- 'div+ table tr:nth-child(2) td:nth-child(4)'
MGD <- 'th~ td+ td, th~ td+ td'

name <- the_durham_url %>% html_nodes(name) %>% html_text()
id <- the_durham_url %>%   html_nodes(id) %>%  html_text()
ownership <- the_durham_url %>% html_nodes(ownership) %>% html_text()
MGD <- the_durham_url %>% html_nodes(MGD) %>% html_text()

df_url <- data.frame("Month" =
                          c("Jan","Feb","Mar","Apr","May","Jun","Jul","Aug","Sept","Oct","Nov","Dec"), 
                             "Year" = rep(year,12),
                             "Maximum_Day_Use" = as.numeric(MGD),
                              "Water_System" = rep(name, 12),
                              "PWSID" = rep(id, 12),
                              "Ownership" = rep(ownership, 12)) %>% 
    mutate(
    Date = my(paste(Month,Year))) %>% 
  arrange(Date)
  return(df_url)
}

```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2020. Then show the structure of the dataframe with either the `str()` or the `glimpse()` function. 

```{r fetch.and.plot.Durham.2020.data}
#7 
df_2020 <- scrape.it(2020, "03-32-010")
str(df_2020)

ggplot(df_2020,aes(x=Date,y=Maximum_Day_Use)) + 
  geom_line() + 
  geom_smooth(method="loess",se=FALSE) +
  labs(
    title = paste(
      "Maximum Day Use (MGD) in 2020"
      ),
    y="Maximum Day Use",
    x="Date"
    )
```

8. Use the function above to extract data for Cary (PWSID = '03-92-020') in 2020. Combine this data with the Durham data collected above and create a plot that compares Cary's to Durham's water withdrawals.

```{r fetch.and.plot.Cary.2020.data}
#8 
df_cary <- scrape.it(2020, "03-92-020")
str(df_cary)

ggplot(df_cary,aes(x=Date,y=Maximum_Day_Use)) + 
  geom_line() + 
  geom_smooth(method="loess",se=FALSE) +
  labs(
    title = paste(
      "Maximum Day Use (MGD) in 2020",
    subsitle = "Cary Municipal LWSP"
      ),
    y="Maximum Day Use",
    x="Date"
    )
str(df_cary)

combine_plots <- bind_rows(df_2020, df_cary)
ggplot(combine_plots,aes(x=Date,y=Maximum_Day_Use,color = Water_System)) + 
  geom_line() + 
  geom_smooth(method="loess",se=FALSE) +
  labs(
    title = "Maximum Daily Withdrawals between Durham and Cary",
    y="Maximum Day Use",
    x="Date"
    )

```


9. Use the code & function you created above to plot Cary's max daily withdrawal by months for the years 2018 thru 2023.Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "10_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bind_rows()` to combine the dataframes into a single one, and use that to construct your plot.

```{r fetch.and.plot.Cary.multiyear.data }
#9 
library(purrr)
library(dplyr)
library(ggplot2)

cary_id <- "03-92-020"
years <- c(2018, 2019, 2020, 2021, 2022, 2023)

the_df <- map2(years, cary_id, scrape.it) 
df_combine <- bind_rows(the_df)


ggplot(df_combine,aes(x=Date,y=Maximum_Day_Use,color=Date)) + 
  #geom_line() + 
  geom_smooth(method="loess",se=FALSE) +
  labs(title = "Cary Water usage data from 2018 to 2023",
       y="Maximum Day Use",
       x="Date")

```

Question: Just by looking at the plot (i.e. not running statistics), does Cary have a trend in water usage over time?
> Answer: There is a clear trend in water usage in Cary where 2020 being the lowest 
> and 2022 being the highest. There is a drop after 2022. 
>
